sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-01-24 doc/paper-jss/inla-comparison.Rmd
Copy editing

We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
\pkg{INLA} requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:

@@

We will start by building identical meshes using both sdmTMB and INLA.
Here, X and Y are coordinates in UTM zone 9.
These same functions are used internally in the sdmTMB `make_mesh()` function.

@@

Here, the meshes should be the same.

@@

Data preparation for \pkg{INLA} models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with TMB.
We use weak priors on the factor fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.

@@

With sdmTMB, we can extract simulated draws from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.

@@

```{r compare-plot, out.width="6in", fig.cap="Blue histograms represent simulated draws from the joint precision matrix of the sdmTMB model compared with the marginal posterior densities from the INLA model.", fig.align='center'}
ggplot(post_inla2) +
  geom_histogram(
    data = post_sdmTMB, aes(.value, after_stat(density)), bins = 30,
    fill = "blue", alpha = 0.2
  ) +
  geom_line(data = post_inla2, aes(x, y), inherit.aes = FALSE) +
  facet_wrap(~variable, scales = "free")
```

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`, equivalent to inverse logit here), to produce occurrence probabilities:

@@

We can get similar predictions with \pkg{inlabru} using only a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation (Fig.\ \ref{fig:inlabru-pred}).

@@

```{r inla-pred, out.width="5.5in"}
# idx <- inla.stack.index(sdat, "stdata")$data
# d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]
# 
# ggplot(d) +
#   geom_point(aes(X, Y, colour = inla_pred), size = 2) +
#   facet_wrap(~year) +
#   scale_colour_viridis_c() +
#   coord_fixed()
```

>>>>

We will use a data set built into the package: trawl survey data for Pacific cod in Queen Charlotte Sound as described in earlier appendices.
Since \pkg{INLA} requires sequentially numbered time steps, we start by selecting the years we will include in this model and creating the year index variable `i_year`:

@@

We will start by building identical meshes using both \pkg{sdmTMB} and \pkg{INLA}.
Here, X and Y are coordinates in UTM zone 9.
These same functions are used internally in the \pkg{sdmTMB} `make_mesh()` function.

@@

These meshes should be the same.

@@

Data preparation for \pkg{INLA} models includes a formula, as well as objects made using several different functions including: `inla.spde.make.index()`, `inla.spde.make.A()`, and `inla.stack()`.
We use `control.inla = list(int.strategy = "eb", strategy = "gaussian")` for comparability with \pkg{TMB}.
We use weak priors on the factor fixed effects (`mean = 0, prec = 1/(100*100)`), which amounts to a `Normal(0, 100^2)` prior.

@@

With \pkg{sdmTMB}, we can extract simulated draws from the joint precision matrix using the `spread_sims()` or `gather_sims()` functions.

@@

```{r compare-plot, out.width="6in", fig.cap="Blue histograms represent simulated draws from the joint precision matrix of the \\pkg{sdmTMB} model compared with the marginal posterior densities from the \\pkg{INLA} model.", fig.align='center'}
ggplot(post_inla2) +
  geom_histogram(
    data = post_sdmTMB, aes(.value, after_stat(density)), bins = 30,
    fill = "blue", alpha = 0.2
  ) +
  geom_line(data = post_inla2, aes(x, y), inherit.aes = FALSE) +
  facet_wrap(~variable, scales = "free")
```

@@

We can predict on a grid (`qcs_grid`) provided with the `pcod` data in the \pkg{sdmTMB} package to map occurrence probabilities in each year in space (Fig.\ \ref{fig:maps}).
For \pkg{sdmTMB}, the resulting values are in link space and require the appropriate inverse-link transformation (always contained in `model$family$linkinv()`---the inverse logit here), to produce occurrence probabilities:

@@

We can get similar predictions with \pkg{inlabru} using a few additional steps: we must recreate `i_year`, create a `SpatialPixelsDataFrame` class data frame, and specify in `formula` to include both the random field (`f`), the factors (`fac`), and the appropriate inverse-link transformation (Fig.\ \ref{fig:inlabru-pred}).

@@

```{r inla-pred, out.width="5.5in"}
idx <- inla.stack.index(sdat, "stdata")$data
d$inla_pred <- m_inla$summary.fitted.values[idx, "mean"]

ggplot(d) +
  geom_point(aes(X, Y, colour = inla_pred), size = 2) +
  facet_wrap(~year) +
  scale_colour_viridis_c() +
  coord_fixed()
```
